Application of Micro-Rotors to Induce Crystal Nucleation: Challenges, Takeaways, and Future Prospects\\ PhD Thesis

### 2025-01-26 Latex_files/Thesis/chapters/chapter 1/main.tex
updates

\chapter{Introduction}
Crystallisation is one of the oldest forms of phase 
separation used by humanity\cite{Schoen1956}, put 
simply, it is the formation and growth of a new 
structured phase within a disordered bulk phase. 
This has applications in a number of industries such 
as pharmaceuticals \cite{Gao2017}, food production 
\cite{Hartel2002}, and electronics \cite{Myerson2002}. 
Where the extraction of dilute materials can help 
improve product quality while keeping production costs 
low. The industrialisation of crystallisation has 
allowed engineers to reliably and efficiently induce 
the crystal formation within a bulk phase. However, 
while on a large scale crystallisation is seemingly an 
understood physical process, just a small amount of 
investigation into the literature reveals that at a 
micro scale there is yet to be unifying theory that can 
accurately explain the process of crystallisation 
\cite{Fu2021}. Crystallisation can be subdivided into 
two steps: Nucleation and Crystal growth. The latter 
focusing on how an already stable crystal grows and how 
it takes on its final shape. Whereas the former is more 
concerned with the first initial moments of a crystals 
formation within a bulk phase.

In this chapter we will outline our current understanding 
of nucleation,why there is a gap in the literature, and why 
there is still a need for local control of nucleation events. 
Furthermore, we will also highlight recent developments 
involving the use of optical tweezers (and other laser based 
methods) in order to develop our understanding of nucleation. 
Lastly we highlight the connection between optical tweezer's 
ability to probe fluid viscosity via rotational motion, and 
the known link between nucleation and fluid flow, to propose 
a novel potential method of creating localised nucleation events.

@@

\subsection{Two Step Nucleation}
The two step nucleation theory is an extension to the CNT that suggests 
that prior to nucleation, the solute will cluster together. This has 
been shown via colloid simulations using protein molecules that have 
significant short range interactions \cite{Wolde1997, Gliko2005}. These 
interactions allow for the formation of a liquid-liquid metastable phase 
from which a new solid phase could form \cite{Anderson2002, Karthika2016}. 
Several papers later reported the presence of stable liquid-like clusters 
that formed prior to nucleation \cite{Savage2009, Wolde1997, Soga1999}. 
The formation of these clusters can be understood by Oswald's rule; which 
says that any crystallising system does not immediately take the path to 
the lowest possible energy state but instead first transitions to the state 
with the smallest free energy barrier \cite{Ostwald1897}. Further phase 
transitions can still occur but this pathway minimises the overall free 
energy cost. Initially it was suspected that the formation of the dense 
liquid phase was a result of stochastic density fluctuations in the system. 
However, simulations suggest that the the local bond order is a stronger 
driving force than the local density \cite{Tan2013}. Studies into two step 
nucleation in Glycine solutions have noted that the solutions need to be 
aged before the presence of clusters was detected \cite{Liao2022}. This 
suggests that there exist further barriers to the formation of clusters 
in some systems, this has lead to instead calling this phenomena as 
Multi-step Nucleation.

\subsection{Non-classical Nucleation} 
The current research into two step nucleation (or now more commonly 
referred to as multi-step nucleation theory) is developing a robust 
framework to describe what nucleation pathway will occur given the 
initial conditions. Reviews of all currently documented nucleation 
pathways highlighted the need for the development of \textit{in situ} 
techniques that can induce nucleation locally but can also reliably 
identify the nucleation pathway across a broad range of experimental 
conditions \cite{Karthika2016, Fu2021}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Crystallisation methods}
\subsection{Cooling Crystallisation}
For some binary mixtures the supersaturation is heavily dependent 
on the solution temperature, therefore a simple method of producing 
crystals is by cooling the mixture to induce crystal formation. At 
ambient temperatures the solute concentration is too high to be 
fully incorporated into the solution ($S\gg 1$), after heating however 
the solute is fully dissolved into the solution ($S<1$). Now as the 
mixture is allowed to cool to room temperature the supersaturation 
will increase until crystal formation begins, the rate of cooling 
drastically influences the size and number of crystals produced.

If $dT/dt$ is high then the final product will consist of large 
crystal and be low in number, as the nucleation rate is directly 
related to the supersaturation only a handful of nuclei can form 
before the remaining solute grows onto the surface. If $dT/dt$ is 
low then the final product will consist of smaller crystals and be 
far more numerous, as the supersaturation is so large that the new 
nuclei are forming continuously. Between these two extremes, one 
can define a meta-stable zone, a region in the temperature-
concentration phase space where both nucleation and crystal growth
can be reliably controlled. The lower limit being given by the 
solubility curve of the binary mixture, and the upper limit being 
defined by the point where crystals begin to spontaneously nucleate
within the solution.

\begin{figure}[h!]
	\includegraphics[width=\linewidth]{MSZW.png}
	\caption{Typical Concentration vs Temperature plot with the 
		solubility curve (black) and metastable zone (red). 3 
		different cooling curves are shown as well: a high rate 
		of cooling (orange) shows the mixture quickly dropping 
		below the solubility curve resulting in no new nuclei 
		forming. A low rate of cooling (green) shows the mixture 
		exceeding the metastable zone, where now nucleation occurs 
		spontaneously and cannot be controlled without further 
		cooling. A typical cooling rate (blue) shows how a 
		typical cooling crystalliser will operate, sitting in the 
		middle of the other two curves.}
	\label{fig:MSZW}
\end{figure}

The viability of cooling crystallisation is dependent on the 
meta-stable zone width, too narrow and the process is difficult to 
control, too wide and the crystal growth rate may be insufficient 
for the desired outcome.

\subsection{Evaporative Crystallisation}
In situations where control of the final product size or shape 
is not the main focus, often the cheapest method of producing a 
crystalline product is simply to allow the solvent evaporate and 
separate from the solute. Depending on the total volume of solvent 
to evaporate this process can take on time scale of several days 
to complete. This is often used for the formation of seed crystals 
when larger seed crystals are required \cite{Srinivasan2008, Li1992}.

\subsection{Anti-solvent Crystallisation} 
Anti-solvent crystallisation is a more involved separation method,
by adding a new solvent that is miscible with the old solvent but 
is immiscible with the solute, the solute's effective solubility 
is decreased. As such, the solution becomes supersaturated. This 
has the trade off in that the overall mass ratio between the solvent 
and solute is now lower. Careful measurement of the anti-solvent is 
required in order to ensure that the decrease in solubility is not 
out matched by the decrease in mass ratio.Anti-solvent crystallisation 
is not as common due to the fact that not all solutions will have an 
ideal anti-solvent candidate.

@@

>>>>

\chapter{Introduction}
Crystallisation is one of the oldest forms of phase 
separation used by humanity \cite{Schoen1956}, put 
simply, it is the formation and growth of a new 
structured phase within a disordered bulk phase. 
This has applications in a number of industries such 
as pharmaceuticals \cite{Gao2017}, food production 
\cite{Hartel2002}, and electronics \cite{Myerson2002}. 
Where the extraction of dilute materials can help 
improve product quality while keeping production costs 
low. The industrialisation of crystallisation has 
allowed engineers to reliably and efficiently induce 
the crystal formation within a bulk phase. However, 
while on a large scale crystallisation is seemingly an 
understood physical process, just a small amount of 
investigation into the literature reveals that at a 
micro scale there is yet to be unifying theory that can 
accurately explain the process of crystallisation 
\cite{Fu2021}. Crystallisation can be subdivided into 
two steps: Nucleation and Crystal growth. The latter 
focusing on how an already stable crystal grows and how 
it takes on its final shape. Whereas the former is more 
concerned with the first initial moments of a crystals 
formation within a bulk phase.

In this chapter we will outline our current understanding 
of nucleation, why there is a gap in the literature, and why 
there is still a need for local control of nucleation events. 
Furthermore, we will also highlight recent developments 
involving the use of optical tweezers (and other laser based 
methods) in order to develop our understanding of nucleation. 
Lastly we highlight the connection between optical tweezer's 
ability to probe fluid viscosity via rotational motion, and 
the known link between nucleation and fluid flow, to propose 
a novel potential method of creating localised nucleation events.

@@

\subsection{Two Step Nucleation}
The two step nucleation theory is an extension to the CNT 
that suggests that prior to nucleation, the solute will 
form precursor structures. The CNT assumes that solute 
structures below the critical radius should be unstable 
within the solution due to interfacial forces trying to 
dissolve the solute. This assumption is based on the idea 
that any interactions between solute molecules are much
smaller than the surface tension due to the bulk solution.
If instead the solute molecule interactions are significant
then it stands to reason that subcritical liquid-like 
structures are stable.

The first observation of a liquid-like structure was in 
$CaCO_3$ solutions, initially it was assumed that polymers
were a necessary additive to create a precursor structure 
\cite{Driessche2017, Karthika2016}. This is not the case when 
the solution is above its critical temperature, allowing the 
solution to be in the binodal region of the phase diagram. 
This would result in the formation of two liquid phases, 
one being rich in solute, the other being solute poor 
\cite{Karthika2016, Fu2021}. Several papers later reported 
the presence of stable liquid-like clusters that formed 
prior to nucleation \cite{Savage2009, Wolde1997, Soga1999}.

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{oswalds_rule.jpg}
	\caption{Free energy diagram of Oswald's rule for a 
	crystal with three possible polymorphs. The diagram 
	shows that there exist local minima in the free energy 
	that represent the different polymorphic forms. A similar
	free energy diagram can be used to describe the two-step
	nucleation mechanism.}
\end{figure}

The formation of these clusters can be understood by Oswald's 
rule; which says that any crystallising system does not 
immediately take the path to the lowest possible energy state 
but instead first transitions to the state with the smallest 
free energy barrier \cite{Ostwald1897}. Oswald's rule was 
assumed to be applicable for crystal polymorphs but can be
generalised to describe the nucleation process also. Further phase 
transitions can still occur but the pathway taken should 
always minimises the overall free energy cost. It is assumed
that between the amorphous phase and the crystalline phase 
the solute undergoes dehydration while changing structurally 
into a more ordered phase, though there is some discussion 
whether the process merely undergoes dehydration, or if there 
is a more complex chemical reaction
\cite{Karthika2016}.

With regards to liquid-precursors, there is still a lot of
discussion about the relationship between the solution 
conditions and the the precursors. The size and rate at 
which these precursors agglomerate based on the degree of 
saturation would provide invaluable information on the 
kinetics of two-step nucleation \cite{Fu2021}. In some 
cases it has even been shown that 'ageing' solutions is 
a necessary step in order to detect the presence of 
stable amorphous clusters \cite{Liao2022}. This was in 
solutions that were supersaturated and yet after the 
formation of precursors spontaneous nucleation was not
seen even after several days of observation \cite{Liao2022}.
This suggests that in some cases the precursors are
metastable and require an external trigger to induce
crystallisation. The presence of metastable structures
and other phenomena has lead to development of a more 
universal theory, namely non-classical nucleation.

\subsection{Non-classical Nucleation} 
Two-step nucleation is based on the idea that between
the bulk solution and the final crystalline phase there
exists and intermediate stage. While this is broadly an
accepted concept, research has shown that in some cases
the intermediate stage can be subdivided into multiple 
states. This has been shown to be the case with metals
in both solid and liquid conditions \cite{Cao2020, Ye2023}
and colloidal simulations have indicated that the 
precursors undergo structural change over time \cite{Tan2013}. 
While in some cases it has been possible to directly
monitor the structural change in the amorphous clusters, 
many organic compounds have not received such treatment
due to the difficulty of characterising and monitoring 
the pre nucleation stage \cite{Fu2021}. This is often 
why two-step nucleation is simply referred to as 
multi-step nucleation.

There are also a myriad of other unique nucleation events
that deviate significantly from both the classical and 
two-step nucleation theories. A review paper discussing
recent advances in non-classical crystallisation highlighted
that depending on the collision kinetics of the system 
and the molecular interactions could result in drastically
different reaction kinetics. In the former case, unstable
subcritical nuclei have been shown to collide and coalesce
into a supercritical nucleus, but only in conditions where 
subcritical nucleation are plentiful despite a low degree
of supersaturation \cite{Baumgartner2013}. The latter case 
is more prevalent for dipole solute molecules, the anisotropic 
attraction results in nuclei with non-spherical geometries, 
this results in varying kinetics for agglomeration dependent 
on the orientation of each molecule \cite{Yau2001}. Phenomena 
such as the above examples has been collectively refereed to 
as non classical nucleation, it is known that they must be 
related to both CNT and multi-step nucleation but the connection 
is not clear based on the current literature.

As such, much of the research into nucleation theory is 
focused on developing \textit{in situ} techniques that can 
monitor and characterise the pre nucleation stage, this
would glean information on cases when both classical 
and non-classical nucleation is a possibility 
\cite{Karthika2016}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{In situ techniques for studying nucleation}
There are several different phenomena \cite{Fu2021, Karthika2016} 
that are not well described by any individual theoretical framework 
(CNT or multi-step nucleation for example). There is an increasing 
interest in probing the pre-nucleated solution.

One of the key conclusions is the need for the development of 
better \textit{in situ} techniques for characterizing and 
imaging the nucleation pathways. Understanding which specific 
pathway is unfolding under a certain set of experimental 
parameters would allow one to develop a map of the thermodynamic 
landscape available to a pre-nucleation system. The challenge 
therein lies in developing experimental methods by which one can 
reproduce similar nucleation events and in turn use 
characterisation techniques to identify the pathway taken. The 
method not only needs to be repeatable fora single set of system 
parameters, but also needs to be flexible enough to replicate 
results regardless of the system parameters (i.e. varying 
supersaturation, temperature, and solute choice). There have 
been several different methods deployed to try and observe 
crystallisation at a microscopic scale.

\subsection{Computer Simulations}
While not a direct observation of nucleation events, computer
simulations have proved an invaluable tool for studying and 
testing predictions about the pre-nucleated solution. By 
comparing the simulative nucleation rate to the expected 
rate given by current models, researchers can test newly 
developed theories and glean information about the microscopic
parameters of a system.

The main challenge facing computer simulations is the issue
of limited scope. Molecular dynamic simulations are constrained
by their choice of time and length scale, for example simulations
in a canonical ensemble face issues where the chemical potential
is depleted to the point that the driving force for crystallisation
is halted \cite{DuranOlivencia2015, Finney2023}, this can be 
addressed but requires an extensive increase in computational 
resources to maintain the chemical potential equilibria. Like wise
in most cases the choice of time scale is a crucial factor, shorter
time scales provide a more accurate evolution of the solution; 
however, in some cases the time scale magnitude for crystallisation
is often far greater than the time scale used in the simulation. This
has led to more advanced sampling techniques that allow for longer 
time scales which has provided insights into the pre-nucleation 
solution stage and formation of pre-nucleated clusters \cite{Finney2023}.

The study of the pre-nucleated solution and the evolution of 
precursors has been aided greatly by the use of simulative 
studies. Early on it was shown in colloidal crystallisation 
that the structure of the precursor was time dependent; 
colloidal simulations showed that regardless of the molecular
interactions between colloids the atomic packing pattern was 
predominately hexagonal centred, over time the packing
became dominated by body centred packing in the case 
where the interactions were longer reaching \cite{Tan2013}. 
They also confirmed that precursor formation was driven not
by local density variations, but instead was driven by the 
local bond order of the colloidal precursors \cite{Tan2013}, 
providing strong evidence that the colloidal pre-nucleation 
clusters undergo rearrangement prior to nucleation. Further
developments in $NaCl$ simulations have shown that by varying
the density fluctuations the precursor structure could either
be amorphous, the classical rock-salt structure, or a new 
structure similar to wurtzite. While this result is difficult
to confirm it provides a important conclusion, that simulations
can hint at new nucleation pathways that have not been confirmed
experimentally. This means that computer simulations can not
be merely used to confirm theories, but can be used to examine
and search for novel results.

Overall computer simulations make up a core area of research 
in the study of nucleation. The technique is however limited,
often struggling to examine nucleation over longer time scales
or larger number of particles. While sampling methods and 
enhanced simulation techniques has alleviated this somewhat 
they are still far from being able to describe the behaviour 
of larger molecules undergoing nucleation.

\subsection{Transmission Electron Microscope (TEM)}
One of the more well known techniques used for direct observations 
of crystallisation dynamics is using transmission electron 
microscopy (TEM). The basic working principle involves creating a 
focused beam of electrons that are directed onto a target sample.
Due to the wave-particle duality, the electrons can be treated 
as a unified beam of light, whose wavelength is dependent on the 
kinetic energy of the electrons. The wavelength of an individual 
electron is given by \cite{Williams2009}:
\begin{align}
	\lambda_e = \frac{h}{\sqrt{2mE(1+\frac{E}{2mc^2})}}
\end{align}

\noindent
When incident on a sample the wave is scattered by the sample 
and can then be detected by an imaging screen that measures 
the difference in intensity between the scattered and 
transmitted beams. The advantage of TEM is that one can bypass
the diffraction limit imposed by visible light, the maximum 
resolution that can be achieved by a microscope is given by
\cite{Champness2020}:
\begin{align}
	d = \frac{\lambda}{2NA}
\end{align}

\noindent
where NA is the numerical aperture of the objective lens, and
$\lambda$ is the wavelength of light used to illuminate a sample.
For a visible light LED with an average wavelength of $550~nm$, 
the upper limit for an objective lens (NA of 1.2) is around 
$230~nm$ whereas a similar lens focusing an electron beam 
with an average kinetic energy of $100~keV$ will have a maximum 
resolution of just over $4~pm$ \cite{Champness2020}. There is
an upper limit to the maximum resolution possible as electron 
lens are far inferior to what modern optics can achieve 
\cite{Champness2020} so there is a trade off between quality and 
resolution.

TEM technology has proven invaluable to the study of nucleation
at an atomic level. The samples are often only a few hundred 
atoms thick in order to optimise the imaging results. Much of 
the supporting evidence for non-classical nucleation has been 
demonstrated via TEM \cite{Ye2023}. For example when Cao 
\textit{et al} observed the formation of amorphous atomic 
cluster using $\gamma-$Fe,Au, and Re; all three of which 
underwent continuous transformation before eventually 
reaching an final ordered state \cite{Cao2020}. The amorphous 
cluster did not immediately transition to an ordered structure, 
instead fluctuating somewhat randomly until eventually settling 
into a crystalline structure \cite{Cao2020}. This was confirmed 
both in solid growth and liquid growth conditions \cite{Cao2020, 
Loh2016}, suggesting that the non-classical nucleation path 
dominates in all cases of metal crystal formation \cite{Ye2023}. 
Confirmation is necessary for different solute choices, as in 
the case of organic solids it is believed that non-classical 
nucleation only dominates in low supersaturations whereas 
classical nucleation is dominate at higher supersaturations.

TEM analysis has not been extended to the study of organic 
crystals, proteins, or simple salts at an atomic level 
however. The main reason being that organic materials can be 
damaged by the electron beam and therefore need to be frozen 
prior to being imaged \cite{Champness2020}. This was used to 
study the polymorphic time dependence of glycine crystals, by 
varying the time prior to freezing Broadhurst \textit{et al} 
showed that $\beta$ glycine dominates initially but $\alpha$-
glycine is more stable after a longer period, $\gamma$-glycine 
was only found when allowed to dry over an hour on a glass slide \cite{Broadhurst2020}. While this was invaluable for showing 
the time scales with which polymorphs form, it doesn't provide 
much information on the kinetics of crystal growth as these 
were snap shots and not direct observations of the nucleation 
process.

Overall TEM analysis is invaluable for the study of atomic 
crystal growth, the secondary scattering from electron 
beams also allows for chemical and structural information 
to be gleaned from the target. This has given strong evidence
for the non-classical nucleation theory, particularly in 
metal crystal formation. Due to the high energy of the 
electrons, organic materials have not received the same
focus outside of cryoTEM analysis. Non-invasive methods
are therefore necessary to gain an understanding of the
crystal growth and formation of organic crystals, 
particularly when both classical and multi-step nucleation 
can occur.

@@

\subsection{Characterisation of rotational motion}

Rotational motion about a single axis is easiest to account for.  
When the power spectra of elliptical polystyrene particles was 
fitted by Yogesh \textit{et al} \cite{Yogesha2011PreciseCO}, they 
assumed that the rotational motion was purely in the transverse
plane. As such they did not have to account for any variance in 
the trapping strength due to orientation nor did they need to
consider non-periodic rotational behaviour. In the case where 
rotational motion is stochastic the problem is more complex. 
For example, when an optical fibre trap characterisation 
technique was implemented by Saffron \textit{et al} 
\cite{BarZiv1997, Meller1998}, they were able to use dynamic 
light scattering to characterise both the axial and lateral 
trap stiffness acting on microspheres. The only drawback 
admitted to in their work was that the technique was constrained 
to isotropic scatters as their theoretical model for describing 
the auto-correlation function was predicated on the fact that 
any variations in the signal are due to the particles 
translational motion within the confines of a cylindrical trap 
\cite{BarZiv1997}. Where the upper limit of the cylindrical 
trap is given by the Rayleigh range ($z_R = n\omega_0/NA$).
However, as demonstrated by the results from Chapter 4.1, 
the axial traps of spherical aggregates is often situated 
far beyond the Rayleigh range (for a 1.2 NA laser this is 
$\pm5.985 \mu m$).
